Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-04 manuscript/manuscript.tex
edits with GC

 Rapid phenotypic adaptation is often observed in natural populations and
 selection experiments.  However, detecting the impact of this selection
 genome-wide is difficult, since adaptation often proceeds from standing
 variation and selection on highly polygenic traits, both of which may leave
 faint genomic signals indistinguishable from a noisy background of genetic
 drift.  One promising signal comes from the genome-wide covariance between
 allele frequency changes observable from temporal genomic data, e.g.
 evolve-and-resequence studies. These temporal covariances reflect how the
 change in neutral allele frequency at one timepoint is predictive of the
 changes at later timepoints when there is heritable fitness variation in the
 population, as neutral alleles can remain associated with selected alleles
 over time. Since genetic drift does not lead to temporal covariance, we can
 use these covariances to estimate what fraction of the variation in allele
 frequency change through time is driven by linked selection. Here, we
 reanalyze two \emph{Drosophila simulans} evolve-and-resequence studies, and
 one artificial selection experiment in mice, to quantify the effects of linked
 selection over short timescales using covariance among time-points and across
 replicates. We estimate that over 20\% of allele frequency change in these
 experiments is driven by selection in some experiments. Against this
 background of positive genome-wide temporal covariances we also identify
 signals of negative temporal covariance corresponding to reversals in the
 direction of selection for a reasonable proportion of loci over the time
 course of a selection experiment.  Overall, we find that in the three studies
 we analyzed, linked selection has a large impact on short-term allele
 frequency dynamics that is readily distinguishable from genetic drift.

@@

\vb{One concern is that these covariances reflect a few large-effect loci
  rather than selection on a polygenic trait. Since our covariances are
  essentially averages over loci, the covariance estimate can be strongly
  affected by a single outlier region. To test whether large outlier regions
  drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
  calculate the covariances in 100kb windows along the genome (we refer to
  these as windowed covariances throughout) and exclude the upper and lower
  $\alpha$ tails to calculate the trimmed average windowed covariance. As
  $\alpha$ increases to 0.5, the trimmed covariance converges to the median
  windowed covariance (by the definition of the median; see Supplementary
  Figure \ref{suppfig:barghi-trimmed-mean}). In other words, this median
  windowed covariance is a measure of the genome-wide covariance robust to
  large-effect loci; we find that the median windowed covariance $\cov(\Delta
  p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
  from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]). 
 }

@@

@@

A benefit of between-replicate covariances is that unlike temporal covariances,
these can be calculated with only two sequenced timepoints and a replicated
study design. This allowed us to assess the impact of linked selection in
driving convergent patterns of allele frequency change across replicate
populations in two other studies. First, we reanalyzed the selection experiment
of \textcite{Kelly2019-dc}, which evolved three replicate wild populations of
\emph{Drosophila simulans} for 14 generations adapting to a novel laboratory
environment. Since each replicate was exposed to the same selection pressure
and share linkage disequilibria common to the original natural founding
population, we expected each of the three replicate populations to have
positive between-replicate covariances. We find all three pairwise
between-replicate covariances are positive and statistically significant (95\%
CIs, [0.0063, 0.0086], [0.0064, 0.0086], [0.0061, 0.0083]). We estimate the
convergent correlation coefficient across these replicates as 0.36 ($95\%$
block-bootstrap confidence interval $[0.31, 0.40]$).

@@

\vb{The presence of an unselected control line provides an alternative way to
  partition the effects of linked selection and genetic drift.

@@

A major finding in the Longshanks study were two major-effect loci that showed
parallel frequency shifts between the two selection lines: a region harboring
the gene Nkx3-2 known to be involved in limb development, and another region
harboring six other candidate genes. We were curious to what extent our
genome-wide covariances were being driven by these two outlier large-effect
loci, so we excluded them from the analysis. Since we do not know the extent to
which linkage disequilibrium around these large-effect loci affects neighboring
loci, we took the conservative precaution of excluding the entire chromosomes
these loci reside on (chromosomes 5 and 10), and re-calculating the temporal
covariances. We find excluding these large effect loci has little impact on the
confidence intervals (blue confidence intervals in Figure \ref{fig:figure-2}),
indicating that these across-replicate covariances are indeed driven by a
polygenic signal.

@@

Since the seminal analysis of \textcite{Maynard_Smith1974-lc} demonstrating
that linked neutral diversity is reduced as an advantageous polymorphism arises
and sweeps to fixation, over four decades of theoretical and empirical research
has enriched our understanding of linked selection.  One under-used approach to
understand the genome-wide effects of selection on standing variation, e.g.
selection on an infinitesimal polygenic trait, stems from an early quantitative
genetic model of linked selection \parencite{Robertson1961-ho} and its later
developments (\cite{Santiago1995-hx,Santiago1998-bs}; see also
\cite{Barton2000-zg} for a comparison of these models with classic hitchhiking
models). Implicit in these models is that autocovariance between allele
frequency change is created when there is heritable fitness variation in the
population, a signal that may be readily detected from temporal genomic data
\parencite{Buffalo2019-io}.  Depending on how many loci affect fitness, such a
strong effect of linked selection may not be differentiable from genetic drift
using only single contemporary population samples or looking at at temporal
allele frequency change at each locus in isolation.  In this way, averaging
summaries of temporal data allows us to sidestep the key problem of detecting
selection from standing variation: that the genomic footprint leaves too soft
of a signature to differentiate from a background of genetic drift. In fact we
find that the temporal covariance signal is detectable even in the most
extremely difficult to detect soft sweep case: polygenic selection on highly
polygenic traits \parencite{Buffalo2019-io}.

@@

We would like to thank the authors of the original studies we've analyzed,
including Neda Barghi, Christian Schl{\"o}tterer, John Kelly, Kimberly Hughes,
Frank Chan, Campbell Rolian, Nick Barton, Alan Bergland, and Dmitri Petrov.  We
would also like to thank Doc Edge for helpful statistical advice, and Matt
Osmond for helpful discussions.

@@

To infer the uncertainty of covariance, convergence correlation, and $G(t)$
estimates, we used a block bootstrap procedure. This is a version of the
bootstrap that resamples blocks of data points, rather than individual data
points, to infer the uncertainty of an statistic in the presence of unknown
correlation structure between data. With genome-wide data, linkage
disequilibria between sites creates complex and unknown dependencies between
variants. The estimators used in this paper are predominantly ratios, e.g.
temporal-replicate covariance standardized by half the heterozygosity,
$G(t)$ which is the ratio of covariance to total variance, and the
convergence correlation (equation \eqref{eq:conv-corr}). In these cases, we can
exploit the linearity of the expectation to make the bootstrap procedure more
computationally efficient, by pre-calculating the statistics of the ratio's
numerator and denominator, $N(\mathbf{x}_i)$ and $D(\mathbf{x}_i)$, on the data
$\mathbf{x}_i$ for all blocks $i \in \{1, 2, \ldots, W\}$ in the genome. Then
we draw $W$ bootstrap samples with replacement, and compute the estimate for
bootstrap sample $b$ with an average weighted by the number of loci in all
sampled blocks,

\begin{align}
  \tilde{\theta}_b = \sum_{i=1}^W w_i \frac{N(\mathbf{x}_i)}{D(\mathbf{x}_i)}
\end{align}
%
Note that computing the ratio of averages rather than the average of a ratio is
a practice common for population genetic statistics like $F_{ST}$
\parencite{Bhatia2013-zy}. With these $B$ bootstrap estimates, we calculate the
$\nicefrac{\alpha}{2}$ and $1-\nicefrac{\alpha}{2}$ quantiles, which we use to
estimate the $1-\alpha = 95\%$ pivot confidence intervals (p. 33
\cite{Wasserman2006-jl}, p. 194 \cite{Davison2013-oy}) throughout the paper,

@@

\subsection{Proportion of Variance Explained by Linked Selection across Replicates}

@@

where $\Delta_{_N} p_t$ and $\Delta_{_M} p_t$ are the allele frequency changes
due to non-heritable variation in fitness (e.g. environmental differences in
family size) and Mendelian segregation. In \textcite{Buffalo2019-io} allele
frequency changes due to heritable fitness variation are captured in the term
$\Delta_{_H} p_t$; here we further partition this into heritable frequency
changes shared between replicates ($\Delta_{_S} p_t$) and those unique to a
replicate ($\Delta_{_H} p_t$). By definition,

>>>>

 Rapid phenotypic adaptation is often observed in natural populations and
 selection experiments.  However, detecting the impact of this selection
 genome-wide is difficult, since adaptation often proceeds from standing
 variation and selection on highly polygenic traits, both of which may leave
 faint genomic signals indistinguishable from a noisy background of genetic
 drift.  One promising signal comes from the genome-wide covariance between
 allele frequency changes observable from temporal genomic data, e.g.
 evolve-and-resequence studies. These temporal covariances reflect how the
 change in neutral allele frequency at one timepoint is predictive of the
 changes at later timepoints when there is heritable fitness variation in the
 population, as neutral alleles can remain associated with selected alleles
 over time. Since genetic drift does not lead to temporal covariance, we can
 use these covariances to estimate what fraction of the variation in allele
 frequency change through time is driven by linked selection. Here, we
 reanalyze two \emph{Drosophila simulans} evolve-and-resequence studies, and
 one artificial selection experiment in mice, to quantify the effects of linked
 selection over short timescales using covariance among time-points and across
 replicates. We estimate that over \vb{TO FIX PUT RANGE XXX} 20\% of allele
 frequency change in these experiments is driven by selection in some
 experiments. Against this background of positive genome-wide temporal
 covariances we also identify signals of negative temporal covariance
 corresponding to reversals in the direction of selection for a reasonable
 proportion of loci over the time course of a selection experiment.  Overall,
 we find that in the three studies we analyzed, linked selection has a large
 impact on short-term allele frequency dynamics that is readily distinguishable
 from genetic drift.

@@

\vb{One concern is that these covariances reflect a few large-effect loci
  rather than selection on a polygenic trait. Since our covariances are
  essentially averages over loci, the covariance estimate can be strongly
  affected by a single outlier region. To test whether large outlier regions
  drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
  calculate the covariances in 100kb windows along the genome (we refer to
  these as windowed covariances throughout) and exclude the upper and lower
  $\alpha$ tails to calculate the trimmed average windowed covariance. As
  $\alpha$ increases to 0.5, the trimmed covariance converges to the median
  windowed covariance (by the definition of the median; see Supplementary
  Figure \ref{suppfig:barghi-trimmed-mean}). In other words, this median
  windowed covariance is a measure of the genome-wide covariance robust to
  large-effect loci; we find that the median windowed covariance $\cov(\Delta
  p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
  from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]).

  XXX put a table median for all timepoints.
 }

@@

\vb{We note that blah blha lbah, have a look at the pairwiwe covariances for
Fig 2A}

@@

A benefit of between-replicate covariances is that unlike temporal covariances,
these can be calculated with only two sequenced timepoints and a replicated
study design. This allowed us to assess the impact of linked selection in
driving convergent patterns of allele frequency change across replicate
populations in two other studies. First, we reanalyzed the selection experiment
of \textcite{Kelly2019-dc}, which evolved three replicate wild populations of
\emph{Drosophila simulans} for 14 generations adapting to a novel laboratory
environment. Since each replicate was exposed to the same selection pressure
and share linkage disequilibria common to the original natural founding
population, we expected each of the three replicate populations to have
positive between-replicate covariances. We find all three pairwise
between-replicate covariances are positive and statistically significant (95\%
CIs, [0.0063, 0.0086], [0.0064, 0.0086], [0.0061, 0.0083]). We estimate the
convergent correlation coefficient across these replicates as 0.36 ($95\%$
block-bootstrap confidence interval $[0.31, 0.40]$). \vb{Put in a note about
replicate G.}

@@

\vb{The presence of an unselected control line provides an alternative way to
  partition the effects of linked selection and genetic drift: we can compare
  the total

@@

A major finding in the Longshanks study were two major-effect loci that showed
parallel frequency shifts between the two selection lines: a region harboring
the gene Nkx3-2 known to be involved in limb development, and another region
harboring six other candidate genes. We were curious to what extent our
genome-wide covariances were being driven by these two outlier large-effect
loci, so we excluded them from the analysis. Since we do not know the extent to
which linkage disequilibrium around these large-effect loci affects neighboring
loci, we took the conservative precaution of excluding the entire chromosomes
these loci reside on (chromosomes 5 and 10), and re-calculating the temporal
covariances. We find excluding these large effect loci has little impact on the
confidence intervals (blue confidence intervals in Figure \ref{fig:figure-2}),
indicating that these across-replicate covariances are indeed driven by a large
number of loci. \vb{This is consistent with a signal of selection on a
  polygenic trait driving genome-wide change, although we note that
  large-effect loci can contribute to the indirect change at unlinked loci
\parencite{Robertson1961-ho,Santiago1995-hx}. }

@@

Since the seminal analysis of \textcite{Maynard_Smith1974-lc} demonstrating
that linked neutral diversity is reduced as an advantageous polymorphism arises
and sweeps to fixation, over four decades of theoretical and empirical research
has enriched our understanding of linked selection.  One under-used approach to
understand the genome-wide effects of selection on standing variation, e.g.
selection on an infinitesimal polygenic trait, stems from an early quantitative
genetic model of linked selection \parencite{Robertson1961-ho} and its later
developments
(\cite{Santiago1995-hx,Santiago1998-bs,Wray1990-zf,J_A_Woolliams_N_R_Wray_R_Thompson2008-qo};
see also \cite{Barton2000-zg} for a comparison of these models with classic
hitchhiking models). Implicit in these models is that autocovariance between
allele frequency change is created when there is heritable fitness variation in
the population, a signal that may be readily detected from temporal genomic
data \parencite{Buffalo2019-io}.  Depending on how many loci affect fitness,
such a strong effect of linked selection may not be differentiable from genetic
drift using only single contemporary population samples or looking at at
temporal allele frequency change at each locus in isolation.  In this way,
averaging summaries of temporal data allows us to sidestep the key problem of
detecting selection from standing variation: that the genomic footprint leaves
too soft of a signature to differentiate from a background of genetic drift. In
fact we find that the temporal covariance signal is detectable even in the most
extremely difficult to detect soft sweep case: polygenic selection on highly
polygenic traits \parencite{Buffalo2019-io}.

@@

We would like to thank the authors of the original studies we've analyzed,
including Neda Barghi, Christian Schl{\"o}tterer, John Kelly, Kimberly Hughes,
Frank Chan, Campbell Rolian, Nick Barton, Alan Bergland, and Dmitri Petrov.  We
would also like to thank Doc Edge for helpful statistical advice, and Matt
Osmond for helpful discussions. This research was supported by an NSF Graduate
Research Fellowship grant awarded to VB (1650042), and NIH (R01-GM108779) and
NSF (1353380) awarded to GC.

@@

To infer the uncertainty of covariance, convergence correlation, and $G(t)$
estimates, we used a block bootstrap procedure. This is a version of the
bootstrap that resamples blocks of data points, rather than individual data
points, to infer the uncertainty of an statistic in the presence of unknown
correlation structure between data. With genome-wide data, linkage
disequilibria between sites creates complex and unknown dependencies between
variants. The estimators used in this paper are predominantly ratios, e.g.
temporal-replicate covariance standardized by half the heterozygosity, $G(t)$
which is the ratio of covariance to total variance, and the convergence
correlation (equation \eqref{eq:conv-corr}). In these cases, we can exploit the
linearity of the expectation to make the bootstrap procedure more
computationally efficient, by pre-calculating the statistics of the ratio's
numerator and denominator, $N(\mathbf{x}_i)$ and $D(\mathbf{x}_i)$, on the data
$\mathbf{x}_i$ for all blocks $i \in \{1, 2, \ldots, W\}$ in the genome. Then
we draw $W$ bootstrap samples with replacement, and compute the estimate for
bootstrap sample $b$ with an average weighted by the fraction $w_i$ of total
loci contained in each block,

\begin{align}
  \tilde{\theta}_b = \frac{\sum_{i=1}^W w_i N(\mathbf{x}_i)}{\sum_{i=1}^W w_i D(\mathbf{x}_i)}
\end{align}
%
Note that computing the ratio of averages rather than the average of a ratio is
a practice common for population genetic statistics like $F_{ST}$
\parencite{Bhatia2013-zy}. With these $B$ bootstrap estimates, we calculate the
$\nicefrac{\alpha}{2}$ and $1-\nicefrac{\alpha}{2}$ quantiles, which we use to
estimate the $1-\alpha = 95\%$ pivot confidence intervals (p. 33
\cite{Wasserman2006-jl}, p. 194 \cite{Davison2013-oy}) throughout the paper,

@@

\subsection{Across-Replicate Covariances}

Here, we extend our theoretic work in \textcite{Buffalo2019-io} to
between-replicate allele frequency change covariances. We focus on the
between-replicate covariance at a focal neutral site across two replicate
populations, $A$ and $B$. As in \textcite{Buffalo2019-io}, the heritable allele
frequency change (which is the only term that contributes to the covariance),
can be written as a linear combination of effect sizes $\alpha_{t,l,A}$ and LD
between the neutral site and selected site $l$, $D_{t,l,A}$.

\begin{align}
  \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov\left(\sum_l \alpha_{t,A,l} D_{t,A,l}, \sum_l \alpha_{t,B,l} D_{t,B,l}\right)
\end{align}

Since the replicate populations derive from a common population at time $t=0$
before their split, they share a fraction of their LD plus a replicate-specific
deviation due to replicate-specific selection and drift. Thus, we can partition
the $D$ terms as: $D_{t,A,l} = D_{0,l} + \varepsilon_{t,A}$ (and likewise for
replicate $B$). Under drift alone, this deviation can be found from equation
(4) of \textcite{Ragsdale2019-fs}.  Similarly, in the original population the
additive effects at each locus were identical ($\alpha_{0,l}$), but in each
replicate population have since been perturbed in a replicate-specific manner
(the term $a_{t,A,l}$) as well as a change induced by convergent selection
pressure after $t$ generations, $b_{t,l}$, $\alpha_{0,l} + b_{t,l} + a_{t,A,l}$
(and likewise for replicate $B$).

We then can write,

\begin{align}
  \cov(\Delta p_{t,A}, \Delta p_{t,B}) = \cov\left(\sum_l \alpha_{t,A,l} (D_{0,l} + \varepsilon_{A,l}), \sum_l \alpha_{t,B,l} (D_{0,l} + \varepsilon_{B,l})\right)
\end{align}

@@

where $\Delta_{_N} p_t$ and $\Delta_{_M} p_t$ are the allele frequency changes
due to non-heritable variation in fitness (e.g. environmental differences in
family size) and Mendelian segregation. In \textcite{Buffalo2019-io} allele
frequency changes due to heritable fitness variation are captured in the term
$\Delta_{_H} p_t$; here we further partition this into heritable frequency
changes shared between replicates ($\Delta_{_S} p_t$) and those unique to a
replicate ($\Delta_{_H} p_t$). This is important, as we can

We base this partition on our model in
\textcite{Buffalo2019-io}, where we decompose the LD between the 
focal neutral site and selected sites conditioning on whether

\begin{align}
  \Delta_{_H} p_{t,A} &= \sum_l \alpha_{t,l,A} D_{t,l, A} \\
                      &= \alpha_{t,l,A} D_{t,l,A} + (\alpha_{t,l,A'} + \alpha_{t,l,S}) D_{t,l,S} \\
\end{align}

By definition,
